% ZTL Blueprint: formalization map. All \leanok items are proven in the
% repository's Lean corpus with an empty axiom list.

\chapter{Semantics}

\begin{definition}[Values and the mark]\label{def:values}
  \lean{V}\leanok
  Truth values are $T$ and $F$; the third symbol $Z$ is a mark for an
  unverified input, not a truth value.
\end{definition}

\begin{definition}[Generating principle: the lift]\label{def:lift}
  \uses{def:values}\lean{V.lift1, V.lift2}\leanok
  For a classical connective $f$, its ZTL lift returns $T$ iff $f$ is
  forced to $T$ under \emph{every} classical reading of $Z$; otherwise
  $F$. Truth is never granted on credit.
\end{definition}

\begin{definition}[Connectives]\label{def:ops}
  \uses{def:lift}\lean{V.znot, V.zand, V.zor, V.zimp, V.zxor, V.zxnor}\leanok
  All six connectives are lifts of their classical cores; no table is
  postulated.
\end{definition}

\begin{theorem}[Anchor cells]\label{thm:anchors}
  \uses{def:ops}\lean{V.ax_not_Z, V.ax_notnot_Z, V.ax_xnor_ZZ, V.ax_xor_ZT}\leanok
  The design anchor cells hold: $\neg Z=F$, $\neg\neg Z=T$,
  $Z\leftrightarrow Z=F$, $Z\oplus T=F$, etc.
\end{theorem}

\begin{theorem}[Greediness]\label{thm:greedy}
  \uses{def:ops}\lean{V.evalF_classical}\leanok
  Every compound formula is classical under every valuation: $Z$ lives
  only on atoms.
\end{theorem}

\chapter{Laws and rules}

\begin{theorem}[Modus ponens]\label{thm:mp}
  \uses{def:ops}\lean{V.modus_ponens}\leanok
  From $p=T$ and $p\to q=T$ follows $q=T$: earned truth is transported.
\end{theorem}

\begin{theorem}[Fallen laws]\label{thm:fallen}
  \uses{def:ops}\lean{V.lem_fails, V.deMorgan1_fails, V.contraposition_fails, V.imp_refl_fails}\leanok
  Excluded middle, De Morgan, contraposition-as-identity and $p\to p$
  fail --- all only on $Z$, all are ``truth from form''.
\end{theorem}

\begin{theorem}[Rules vs laws split]\label{thm:split}
  \uses{thm:mp, thm:fallen}\lean{V.rule_contraposition, V.rule_dn_elim_fails}\leanok
  Contraposition survives as an inference rule while failing as an
  identity; double-negation elimination fails even as a rule. The
  deduction theorem holds only left-to-right.
\end{theorem}

\begin{theorem}[No gluts]\label{thm:nogluts}
  \uses{def:ops}\lean{V.no_gluts}\leanok
  $v(p)=v(\neg p)=T$ is unsatisfiable: ZTL is paracomplete, not
  paraconsistent.
\end{theorem}

\chapter{Tableau calculus, certified}

\begin{definition}[Signs]\label{def:signs}
  \uses{def:values}\lean{V.SignT, V.SignF, V.SignP, V.SignN}\leanok
  Strict signs $T,F$ and weak signs $P=\{T,Z\}$, $N=\{F,Z\}$; weak signs
  occur only in refutation polarity.
\end{definition}

\begin{definition}[Engine]\label{def:closes}
  \uses{def:signs, thm:greedy}\lean{V.closes}\leanok
  Fuel-indexed structural tableau engine over the generating basis
  $\{\neg,\wedge,\vee\}$; heavy connectives reduce by the living
  identities.
\end{definition}

\begin{theorem}[Soundness and completeness]\label{thm:closesiff}
  \uses{def:closes}\lean{V.closes_iff}\leanok
  The tableau closes iff the signed nodes are unsatisfiable --- for all
  formulas, by induction on weighted size. Zero axioms.
\end{theorem}

\begin{theorem}[Entailment certificate]\label{thm:tproves}
  \uses{thm:closesiff}\lean{V.tproves_iff}\leanok
  $\Gamma\vdash\varphi$ by the engine iff every valuation making all
  premises $T$ makes $\varphi$ $T$.
\end{theorem}

\begin{theorem}[Native engine agrees]\label{thm:agree}
  \uses{thm:tproves}\lean{V.closesN_iff, V.engines_agree}\leanok
  The engine with native signed rules for $\to,\oplus,\leftrightarrow$
  is certified by the same induction and returns identical verdicts.
\end{theorem}

\chapter{Registers and paradoxes}

\begin{theorem}[Liar is homeless]\label{thm:liar}
  \uses{def:ops}\lean{V.liar_homeless}\leanok
  Negation has no fixed point: no value seats the liar; quarantine is
  irremovable.
\end{theorem}

\begin{theorem}[Kleene register]\label{thm:kleene}
  \uses{thm:liar}\lean{V.kleene_and_monotone, V.eager_and_not_monotone, V.liar_kleene_home}\leanok
  The lazy register is monotone and homes the liar ($\neg Z=Z$); the
  eager register is not monotone: grounding and verdicts require two
  different registers.
\end{theorem}

\begin{theorem}[Carousel and revenge]\label{thm:carousel}
  \uses{thm:kleene}\lean{V.carousel_no_model, V.revenge_bullet}\leanok
  The two-sentence cycle has no eager model over all nine value pairs;
  the revenge sentence's content evaluates to $T$ while the sentence
  stays quarantined --- the price is paid explicitly.
\end{theorem}

\begin{theorem}[Dynamics]\label{thm:dynamics}
  \uses{thm:liar}\lean{V.liar_period2, V.carousel_period4, V.curry_homeless, V.yablo3_unique}\leanok
  Liar oscillates with period 2, the carousel with period 4; Curry is
  homeless without using negation; the finite Yablo truncation has a
  unique grounded model.
\end{theorem}

\chapter{Sets with unverified elements}

\begin{definition}[Marked elements]\label{def:el}
  \uses{def:values}\lean{V.El, V.eqAtom}\leanok
  Elements are verified values or marks; the equality atom yields $Z$
  whenever a mark is involved --- even a mark with itself.
\end{definition}

\begin{theorem}[Marks earn no membership]\label{thm:memz}
  \uses{def:el}\lean{V.memZ}\leanok
  $\mathrm{mem}(z, S)=F$ for every mark $z$ and every list $S$:
  membership is an existential fold with strict witnesses.
\end{theorem}

\begin{theorem}[Identity laws fall on marked sets]\label{thm:marked}
  \uses{thm:memz}\lean{V.sub_marked_false, V.seteq_self_marked}\leanok
  A marked set is not provably a subset of itself; reflexivity of set
  equality fails --- inherited from the tables, not postulated.
\end{theorem}

\chapter{The algebraic passport}

\begin{definition}[J-operators]\label{def:jops}
  \uses{def:ops}\lean{V.jT, V.jF}\leanok
  $J_T(p) = p\wedge p$ (fallen idempotence is the truth detector),
  $J_F(p) = \neg p\wedge\neg p$, $J_Z = \mathrm{isZ}$: three exact,
  disjoint value indicators grown from ZTL's own connectives.
\end{definition}

\begin{theorem}[Expressive completeness, unary]\label{thm:unary}
  \uses{def:jops}\lean{V.unary_complete}\leanok
  Every unary external function $V\to\{T,F\}$ is a J-DNF: one theorem
  over all eight target tables (binary case $512/512$ measured in
  Python).
\end{theorem}

\begin{theorem}[Full deduction theorem for E]\label{thm:ddt}
  \uses{def:jops, thm:greedy}\lean{V.eimp_spec, V.ddt_E}\leanok
  The definable external implication
  $E(p,q)=\neg(p\wedge p)\vee(q\wedge q)$ satisfies
  $\Gamma,\varphi \vDash \psi \iff \Gamma \vDash E(\varphi,\psi)$ in
  both directions, over the whole language; the primitive arrow stays
  one-way.
\end{theorem}

\begin{theorem}[Blok--Pigozzi witnesses]\label{thm:bp}
  \uses{def:jops}\lean{V.dEq_spec, V.bp_refl, V.bp_mp, V.bp_cong_xnor, V.truth_equation, V.bp_iv}\leanok
  The same-value detector $\Delta$ and the truth equation
  $p\wedge p \approx \neg(p\wedge\neg p)$ satisfy conditions
  (i)--(iv) of the Blok--Pigozzi characterization on the matrix:
  ZTL is algebraizable (kinship: Bochvar algebras, Bonzio--Pra Baldi
  2024).
\end{theorem}

\begin{theorem}[Not self-extensional]\label{thm:selfext}
  \uses{def:jops}\lean{V.idem_interderivable, V.not_selfextensional}\leanok
  $p \dashv\vDash p\wedge p$, yet $\neg(p\wedge p) \nvDash \neg p$:
  interderivability is not a congruence --- the same failure that
  builds the detectors.
\end{theorem}

\begin{theorem}[Structurality]\label{thm:structural}
  \uses{thm:greedy}\lean{V.evalF_subst, V.entails_structural}\leanok
  The substitution lemma: $\Gamma\vDash\varphi$ implies
  $\sigma\Gamma\vDash\sigma\varphi$; together with reflexivity,
  monotonicity and cut, $\vDash$ is a structural Tarskian consequence
  relation.
\end{theorem}

\begin{theorem}[Craig interpolation]\label{thm:interp}
  \uses{thm:unary}
  If $A \vDash B$, the J-DNF of $A$'s projection onto the shared atoms
  interpolates: an external function of the shared atoms is always a
  formula. Measured totally ($400/400$ and $32/32$); Lean port on the
  roadmap.
\end{theorem}

\begin{theorem}[Cut admissibility]\label{thm:cut}
  \uses{thm:closesiff}\lean{V.cut_admissible, V.weakening_admissible, V.identity_refutable}\leanok
  The tableau engine read bottom-up is a cut-free sequent calculus;
  cut on the covering pair $\{T\},\{F,Z\}$ is admissible (semantic cut
  elimination on top of the engine certificate), weakening admissible,
  identity derivable.
\end{theorem}
